---
title: "Bikeshare"
author: "Bailee Mull"
date: "18 10 2020"
output: html_notebook
---

The goal of this analysis is to be able to analyze bikeshare data

Prepare workspace:

## Load packages

```{r}
library(tidyverse)
library(janitor)
library(readxl)
library(skimr)
library(summarytools)
library(lubridate)
library(tidyr)
library(ggplot2)
library(sp)
library(sf)
library(rgdal)
```

##Data

The data source from capital bikeshare

## Read the data

```{r}
dfa= read_excel("202008-capitalbikeshare-tripdata-newdata.xlsx", sheet =2) %>% clean_names()
```

```{r}
skim(dfa)
```


## Tidy 

Non Needed

## Subsetting to data of interest.

We will get rid of the following variables:

*ride-id


```{r}
dfb=dfa %>% select(-ride_id)
```
## Ceaning

```{r}
dfc=dfb %>% filter(duration>0, duration<1440)
```

```{r}
dfc %>% filter(duration<300)%>% ggplot(aes(duration)) +geom_histogram(binwidth=1)
```

```{r}
dfsorted= dfc %>% dplyr::arrange(duration)
```

## facet_wrap

```{r}
dfb %>% ggplot(aes(x=duration)) + geom_histogram(bins=1000) + facet_wrap(vars(weekday))
```

```{r}
dfc %>% ggplot(aes(x=started_at,y=duration))+ geom_point()
```

```{r}
dfc %>% ggplot(aes(x=hour,y=duration))+ geom_point()
```


```{r}
dfe=dfb %>% filter(mday(started_at)==3) %>% slice_sample(n=100)
```

## Designing an algorithm

Only start and end times

```{r}
dfe %>% select(start=started_at,end=ended_at)
```

Create a long dataset

```{r}
dfe %>% select(start=started_at,end=ended_at) %>% pivot_longer(start:end)
```

Pick good names, then order by time

```{r}
dfe %>% select(start=started_at,end=ended_at) %>% pivot_longer(start:end, names_to="type",values_to="t") %>% arrange(t)
```

```{r}
dfe %>% select(start=started_at,end=ended_at) %>% pivot_longer(start:end, names_to="type" ,values_to="t") %>% arrange(t) %>% mutate(increment=case_when(type=="start"~1,type=="end"~-1))
```

Cummlative sum of that colum

```{r}
dfe %>% select(start=started_at,end=ended_at) %>% pivot_longer(start:end, names_to="type",values_to="t") %>% arrange(t) %>% mutate(increment=case_when(type=="start"~1, type=="end"~-1)) %>% mutate(riders=cumsum(increment)) 
```

```{r}
dfe %>% select(start=started_at,end=ended_at) %>% pivot_longer(start:end, names_to="type",values_to="t") %>% arrange(t) %>% mutate(increment=case_when(type=="start"~1, type=="end"~-1)) %>% mutate(riders=cumsum(increment)) %>% ggplot(aes(t,riders)) + geom_step()
```
One day started on day three and didn't get returned until the fifth of August.

```{r}
dfb %>% filter(mday(started_at)==3) %>% select(start=started_at,end=ended_at) %>% pivot_longer(start:end, names_to="type",values_to="t") %>% arrange(t) %>% mutate(increment=case_when(type=="start"~1, type=="end"~-1)) %>% mutate(riders=cumsum(increment)) %>% ggplot(aes(t,riders)) + geom_step() + scale_x_datetime(limits=as_datetime(c("2020-08-03","2020-08-04")))
```

Whole range

```{r}
dfb %>% select(start=started_at,end=ended_at) %>% pivot_longer(start:end, names_to="type",values_to="t") %>% arrange(t) %>% mutate(increment=case_when(type=="start"~1, type=="end"~-1)) %>% mutate(riders=cumsum(increment)) %>% ggplot(aes(t,riders)) + geom_step()
```

```{r}
dfb %>% filter(month(started_at)==8,month(ended_at)==8) %>% select(start=started_at,end=ended_at) %>% pivot_longer(start:end, names_to="type",values_to="t") %>% arrange(t) %>% mutate(increment=case_when(type=="start"~1, type=="end"~-1)) %>% mutate(riders=cumsum(increment)) %>% ggplot(aes(t,riders)) + geom_step() + facet_wrap(~mday(t),scales= "free_x",ncol= 7)
```

## Separating eletric and docked bikes

```{r}
dfb %>% filter(month(started_at)==8,month(ended_at)==8) %>% select(rideable_type,start=started_at,end=ended_at) %>% pivot_longer(start:end, names_to="type",values_to="t") %>% arrange(t) %>% mutate(increment=case_when(type=="start"~1, type=="end"~-1)) %>% group_by(rideable_type) %>% mutate(riders=cumsum(increment)) %>% filter(mday(t)<=7) %>% ggplot(aes(t,riders,color=rideable_type)) + geom_step() + facet_wrap(~mday(t),scales= "free_x",ncol= 7)
```

## BOx and whisker plot for trip duration based on "hour

```{r}
dfb %>% filter(duration<60) %>% ggplot(aes(x=hour,y=duration,group=hour)) + geom_boxplot()
```

Adjustment to 31 mini graphs

```{r}
dfb %>% filter(month(started_at)==8,month(ended_at)==8) %>% select(start=started_at,end=ended_at) %>% pivot_longer(start:end, names_to="type",values_to="t") %>% arrange(t) %>% mutate(increment=case_when(type=="start"~1, type=="end"~-1)) %>% mutate(riders=cumsum(increment)) %>% ggplot(aes(t,riders)) + geom_step() + facet_wrap(~mday(t),scales= "free_x",ncol= 7) + theme(axis.text.x = element_blank())
```
Another adjustment to 31 mini graphs

```{r}
dfb %>% filter(month(started_at)==8,month(ended_at)==8,mday(started_at)<=14) %>% select(start=started_at,end=ended_at) %>% pivot_longer(start:end, names_to="type",values_to="t") %>% arrange(t) %>% mutate(increment=case_when(type=="start"~1, type=="end"~-1)) %>% mutate(riders=cumsum(increment)) %>% ggplot(aes(t,riders)) + geom_step() + facet_wrap(~mday(t),scales= "free_x",ncol= 7) + theme(axis.text.x = element_blank())
```

Looking at changing day from 03-04 to 09-10

```{r}
dfb %>% filter(mday(started_at)==9) %>% select(start=started_at,end=ended_at) %>% pivot_longer(start:end, names_to="type",values_to="t") %>% arrange(t) %>% mutate(increment=case_when(type=="start"~1, type=="end"~-1)) %>% mutate(riders=cumsum(increment)) %>% ggplot(aes(t,riders)) + geom_step() + scale_x_datetime(limits=as_datetime(c("2020-08-09","2020-08-10")))
```

Changing day again

```{r}
dfb %>% filter(mday(started_at)==2) %>% select(start=started_at,end=ended_at) %>% pivot_longer(start:end, names_to="type",values_to="t") %>% arrange(t) %>% mutate(increment=case_when(type=="start"~1, type=="end"~-1)) %>% mutate(riders=cumsum(increment)) %>% ggplot(aes(t,riders)) + geom_step() + scale_x_datetime(limits=as_datetime(c("2020-08-02","2020-08-03")))
```

## Looking at eletric vs. docke bikes again

```{r}
dfr=dfb %>% filter(month(started_at)==8,month(ended_at)==8) %>% select(rideable_type,start=started_at,end=ended_at) %>% pivot_longer(start:end, names_to="type",values_to="t") %>% arrange(t) %>% mutate(increment=case_when(type=="start"~1, type=="end"~-1)) %>% group_by(rideable_type) %>% mutate(riders=cumsum(increment))
```

```{r}
dfr1=dfb %>% filter(month(started_at)==8,month(ended_at)==8) %>% select(start=started_at,end=ended_at) %>% pivot_longer(start:end, names_to="type",values_to="t") %>% arrange(t) %>% mutate(increment=case_when(type=="start"~1, type=="end"~-1)) %>% mutate(riders=cumsum(increment))
```

How many riders (max) in any hour time interval?

```{r}
dfrh=dfr1 %>% group_by(hour(t),weekdays(t)) %>% summarise(max_riders=max(riders),ave=mean(riders))
```

## Group 1 mapping steps

```{r}
DC <- readOGR("Preliminary_2020_Census_Tract.csv","Preliminary_2020_Census_Tract.csv". verbose = F)
```




























